Practical Optional Types for Clojure

### 2015-02-27 formal-model.tex
3.4

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable,
allowing us to refine its type.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

@@

\begin{mathpar}
\TIf{}
\end{mathpar}

@@

We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
checking the then branch of the first conditional test is checked in 
an equivalent version of example~\ref{example:decleaf}.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (if (= (get m :op) :if)
      {:op :if, ...}
      (if ...)))
\end{minted}

@@

By T-Abs, checking the body of the function we have the proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
By T-Local we have
$$
\judgement{\propenv{}}{\hastype{m}{\Expr{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{m}
$$
and since {\Expr{}} is a union of HMaps with the entry \makekw{op}, we use T-GetHMap to derive
$$
\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\object{1}}
$$
where \object{1} = {\path{\keype{\makekw{op}}}{m}}. Using our intuitive definition of \equivliteral{} above, we know
$$
\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\object{1}}}{\notprop{\Value{\makekw{if}}}{\object{1}}}}{\emptyobject{}}
$$
Now we check the then branch with our new assumption.
$$
\judgement{\propenvp{}}{\hastype{\curlymap{{\mapvalentrykwkeynoarrow{op}{\makekw{if}}}, ...}}{\Expr{}}}{...}{...}
$$
We are interested in what L-Update can derive in our new environment,
\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.

With the premises
\inpropenv{\propenvp{}}{\isprop{\Expr{}}{m}}
and
\inpropenv{\propenvp{}}{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}
we derive
$$
\inpropenv{\propenvp{}}{\isprop{\update{\Expr{}}{\Value{\makekw{if}}}{\keype{\makekw{op}}}}{m}}
$$
which simplifies to 
$$
\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
                                   {\emptyabsent{}}}{m}}
$$

Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\ref{sec:isaformal}.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
 \classtablelookupsyntax{}
 \begin{mathpar}
  \begin{altgrammar}
    \convertjavatypenil{}
  \end{altgrammar}
  \begin{altgrammar}
    \convertjavatypenonnil{}
  \end{altgrammar}
\end{mathpar}
  \begin{mathpar}
    {\TNewStatic}
    {\TFieldStatic}

@@

We extend the syntax with Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
These come without typing or evaluation rules---Typed Clojure 
demands statically resolved Java interactions
in typed code
(as demonstrated in section~\ref{sec:overviewjavainterop}).
We add type hinted syntax for local variables and let-bindings.

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

To obtain non-reflective calls we define our rewrite relation
$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

Type hint inference is modelled by
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$

To demonstrate we consider the expression 
$$
\fieldexp{getParent}{\hinted{\File}{\x{}}}.
$$
where 
$$
\ct = \{ \File \mapsto \{ \ctfieldsliteral \mapsto \{\ctfldentry{getMethod}{\String}\} \} \}
$$

R-FieldElimRefl eliminates a reflective field.
  $$
    \RFieldElimRefl{}
  $$
First the target is rewritten \rewrite {\taenv{}} {\hinted{\File}{\x{}}}{\hinted{\File}{\x{}}}
then we infer the type hint on the new expression \tajudgement {\taenv{}} {\hastype {\hinted{\File}{\x{}}} {\File}}
We call \fieldtypeliteral{} (figure~\ref{main:figure:javatyping})
to determine the exact signature of the field:
\fieldtype{\ct{}}{\File}{getParent}{\File}{\String}
and return the non-reflective field 
$$
\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}.
$$

Now we present the typing rules (figure~\ref{main:figure:javatyping}).
To briefly finish our example, let us assume \propenv{} = {\isprop{\File}{\x{}}},
then
T-FieldStatic converts the target Java type to a non-nilable
type \javatotc {\File} {\File}} and checks the target against it 
\judgement{\propenv{}}{\hastype{\x{}}{\File}}, and finally converts the return Java type 
to a nilable type and \javatotc {\String} {\Union{\Nil{}}{\String}}
to finish with 
$$
\judgementtwo{\propenv{}}{\hastype{\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}}{\Union{\Nil{}}{\String}}}.
$$

The other rules T-MethodStatic and T-NewStatic work similarly, and vary
primarily in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method simply evaluate their
arguments and call the relevant JVM operation---section~\label{sec:metatheory}
states our exact assumptions about each one.

>>>>

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable,
allowing us to refine its type.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover L-Update in more detail in section~\ref{sec:formalpaths}.

@@

\begin{mathpar}
  {\TIf}
\end{mathpar}

@@

\paragraph{Key path element} We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
checking the then branch of the first conditional test is checked in 
an equivalent version of example~\ref{example:decleaf}.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (if (= (get m :op) :if)
      {:op :if, ...}
      (if ...)))
\end{minted}

@@

We start with proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
Since {\Expr{}} is a union of HMaps, each with the entry \makekw{op}, we can use T-GetHMap.
$$
\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
$$
Using our intuitive definition of \equivliteral{} above, we know
$$
\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
$$
Going down the then branch gives us the extended environment
\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.
Using L-Update we can combine what we know about object $m$ and object
{\path{\keype{\makekw{op}}}{m}}
to derive
$$
\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
                                   {\emptyabsent{}}}{m}}
$$

\paragraph{Class path element} Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\ref{sec:isaformal}.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
 \classtablelookupsyntax{}
 \begin{mathpar}
  \begin{altgrammar}
    \convertjavatypenil{}
  \end{altgrammar}
  \begin{altgrammar}
    \convertjavatypenonnil{}
  \end{altgrammar}
\end{mathpar}
  \begin{mathpar}
    {\TNewStatic}

@@

We extend the syntax with type hinted expressions, Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The Java expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in section~\ref{sec:overviewjavainterop}).

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

To obtain non-reflective calls we define our rewrite relation
\rewrite {\taenv{}} {\e{}} {\ep{}}
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

For example,
R-FieldElimRefl emits a non-reflective field if it can
find a field matching the type hint inferred on \ep{}.
  $$
    \RFieldElimRefl{}
  $$

Type hint inference is modelled by
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$

For demonstration purposes let us rewrite
a simple field expression, with the assumptions that
that we have a class \Point{} with a field \getx{} of Java type \Number
and \intaenv {\taenv{}} {\x{}} {\Point{}}.
In rewriting
the expression \fieldexp{\getx}{\x{}},
{\Point{}} is inferred as the type hint of \x{}
and \Number{} is the field type
by \fieldtypeliteral{} (figure~\ref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
$$

Now we present the typing rules for resolved Java interoperability.
T-FieldStatic checks a resolved field expression by ensuring the target has
the correct static type, then returns a nilable type corresponding the Java type.

\begin{mathpar}
    {\TFieldStatic}
\end{mathpar}

To continue our example, let us assume \propenv{} = {\isprop{\Point}{\x{}}}.
T-FieldStatic checks \x{} under the non-nilable type {\Point}
and returns the nilable type {\Union{\Nil}{\Number}}.

The other rules T-MethodStatic and T-NewStatic work similarly (figure~\ref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method (figure~\ref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation---section~\label{sec:metatheory}
states our exact assumptions about each.

### 2015-02-27 overview.tex
3.4

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}

We can use {\clj{defmethod}} to add a method to \clj{path}.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{File} to the function
\clj{(fn [^File x] (.getPath x))}. 
The call 
$$
\clj{(path (File. "dir/a"))}
$$
now dispatches to the \clj{File} method because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [^File x] (.getPath x)) "dir/a")}.
$$

@@

In the full definition for \clj{path} we add a method for \clj{nil}
which uses the equality check provied by \clj{isa?}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}

>>>>

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.

We can use {\clj{defmethod}} to install a method to \clj{path}.
\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. 
We add another method
which maps
\clj{File} to the function
\clj{(fn [^File x] (.getPath x))}
in the dispatch table.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}

After installing both methods, the call 
$$
\clj{(path (File. "dir/a"))}
$$
dispatches to the second method we installed because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [^File x] (.getPath x)) "dir/a")}.
$$

@@

We include the above sequence of definitions as example~\ref{example:rep}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}
